Informing policy via dynamic models: Cholera in Haiti

### 2023-08-22 ms-plos.Rnw
Removed Vacc Sim Figures

This case study investigates the use of dynamic models and spatiotemporal data to inform a policy decision in the context of the cholera outbreak in Haiti, which started in 2010.
We build on a multi-group modeling exercise by Lee et al.~\cite{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.
Model~1 is stochastic and describes cholera at the national level;
Model~2 is deterministic with spatial structure, and includes transmission via contaminated water;
Model~3 is stochastic with spatial structure, and accounts for measured rainfall.
Model~4 has an agent-based construction, featuring considerable mechanistic detail but limited ability to calibrate these details to data.
We focus on Models~1--3, as the strengths and weaknesses of the agent-based modeling approach \cite{tracy18} are outside the scope of this article.

@@

\begin{figure}
<<Plot_Reported_Cases, echo=FALSE, fig.height=3.5>>=

# Plot reported cholera cases, by department. Creates Figure 1.
ggplot(dep_plot_df, aes(x = date, y = cases + 1)) +
  facet_wrap(~dep, nrow = 2, labeller = dep_labeller) +
  geom_line() +
  theme(axis.title.x = element_blank()) +
  ylab('Reported Cases') +
  scale_y_log10(
    labels = scales::trans_format("log10", scales::math_format(10^.x))
  ) +
  scale_x_date(
    date_labels = "'%y",
    breaks = seq.Date(
      from = as.Date("2011-01-01"),
      to = as.Date("2019-01-01"),
      by = '2 years'
    )
  )
@
\caption{\label{CholeraData}
\small
Weekly reported cholera cases in Haiti from October 2010 to January 2019.
}
\end{figure}

@@

One of the most frequently used approaches to calibrate the parameters of a statistical model is maximum likelihood estimation.
In the case of a POMP model, the likelihood function---as described in Eq.~\myeqref{eq:likedef}---is written as $\mathcal{L}(\paramVec) = \int f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big) d\bm{x}_{0:N}$.
The high-dimensional integral in the likelihood, and the potentially non-linear nature of the process model, make it impossible to obtain a closed form expression of the likelihood function of POMP models except in the most basic cases.
The process model that defines Model~1 was developed with the intent of providing a plausible description of cholera dynamics rather than a providing statistically convenient description.
Because of this, a closed form expression of the likelihood of Model~1 is not readily available.

@@

The ability to maximize the likelihood allows for likelihood-based inference, such as performing statistical tests for potential model improvements.
We demonstrate this capability by proposing a linear trend $\transmissionTrend$ in transmission in Eq.~\myeqref{model1:beta}:
\begin{equation}
\label{model1:betat}
\log\transmission(t) = \sum_{j=1}^6 \transmission_s s_j(t) + \transmissionTrend\bar{t},
\end{equation}
where $\bar{t} = \frac{t - (t_N + t_0) / 2}{t_N - (t_N + t_0) / 2}$, so that $\bar{t} \in [-1, 1]$.
The proposal of a linear trend in transmission is a result of observing an apparent decrease in reported cholera infections from 2012-2019 in Fig.~\ref{CholeraData}.
While several factors may contribute to this decrease, one explanation is that case-area targeted interventions (CATIs), which included education sessions, increased monitoring, household decontamination, soap distribution, and water chlorination in infected areas \cite{rebaudet19CATI}, may have substantially reduced cholera transmission over time \cite{rebaudet21}.

@@

\begin{figure}
\centering
<<Beta_trend_Figure, fig.height=2.4, fig.width=3.6, fig.align='center'>>=
ggplot() +
  geom_point(data = h1_prof_res, aes(x = betat, y = logLik)) +
  geom_line(data = mcap_results$fit, aes(x = parameter, y = smoothed), col = 'blue') +
  geom_vline(xintercept = mcap_results$ci[1], linetype = 'dashed') +
  geom_vline(xintercept = mcap_results$ci[2], linetype = 'dashed') +
  geom_vline(xintercept = mcap_results$mle, col = 'blue') +
  labs(x = "Linear Trend in Transmission", y = 'Log Likelihood') +
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 10))
@
\caption{\label{fig:betat}
Monte Carlo adjusted profile of $\transmissionTrend$.
The blue curve is the profile, the blue line indicates the MLE, and the dashed lines indicate the $95\%$ confidence interval.
}
\end{figure}

We implemented Model~1 using the \code{pomp} package \cite{king16}, relying heavily on the \code{pomp} source code provided by Lee et al.~\cite{lee20}.
Both analyses used the \code{mif2} implementation of the IF2 algorithm to estimate $\paramVec$ by maximum likelihood.
One change we made in the statistical analysis that led to larger model likelihoods was increasing the computational effort in the numerical maximization.
While IF2 enables parameter estimation for a large class of models, the theoretic ability to maximize the likelihood depends on asymptotics in both the number of particles and the number of filtering iterations.
Many Monte Carlo replications are then required to quantify and further reduce the error.
The large increase in the log-likelihood for Model~1 (see Table~\ref{tab:likes}) can primarily be attributed to increasing the computational effort in fitting the model; this result highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.

\begin{figure}
\centering
<<Model_1_Sims_Figure, fig.height=2.35, fig.width=4.6, fig.align='center'>>=
# This code chunk creates the figure containing simulations from Model 1

@@

@@

At the time Lee et al.~\ref{lee20} was published, there was no known algorithm that could readily be used to maximize the likelihood of an arbitrary meta-population POMP model with coupled spatial dynamics, which justifies the spatial decoupling approximation that was used to calibrate model parameters.
For our analysis, we calibrate the parameters of the spatially coupled version of Model~3 using the iterated block particle filter (IBPF) algorithm \cite{ionides22}.
This algorithm extends the work of Ning and Ionides \cite{ning21ibpf}, who provided theoretic justification for the version of the algorithm that only estimates unit-specific parameters.
The IBPF algorithm enables us to directly estimate the parameters of models describing high-dimensional partially-observed nonlinear dynamic systems via likelihood maximization.
The ability to directly estimate parameters of Model~3 is responsible for the large increase in model likelihoods reported in Table~\ref{tab:likes}.
Simulations from the fitted model are displayed in Fig.~\ref{fig:h3spatsims}.

@@

\begin{figure}
<<Model3_SpatPOMP_sims_Figure, fig.height=3>>=

@@

<<Get Model 1 Seasonality>>=

h1_seas_df <- get_h1_seasonality(h1)

@@

\begin{figure}
\centering
<<Model1_Seasonality_Figure, fig.height=2.9, fig.width=5, fig.align='center'>>=
cowplot::plot_grid(gg_rain, gg_trans, align = "v", ncol = 1)
@
\caption{\label{fig:h1SeasRain}
(Top) weekly rainfall in Haiti, lighter colors representing more recent years.
(Bottom) estimated seasonality in the transmission rate (dashed line) plotted alongside mean rainfall (solid line).
The outsided effect of rainfall in the fall may be due to Hurricane Matthew, which struck Haiti in October of 2016 and resulted in an increase of cholera cases in the nation.
}
\end{figure}

@@

Simulations from probabilistic models (Models~1 and~3) represent possible trajectories of the dynamic system under the scientific assumptions of the models.
Estimates of the probability of cholera elimination can therefore be obtained as the proportion of simulations from these models that result in cholera elimination.
The results of these projections are summarized in Figs.~\ref{fig:Mod1Scenarios}--\ref{fig:elimProbs}.

<<Create Model 1 VaccScen plots>>=
gg_m1_V0 <- ggplot() +
  geom_line(data = h1_S0_results$mod1_V0_sims, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_ribbon(data = h1_S0_results$mod1_V0_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  ylab("Log reported cases") +
  theme(axis.title.x = element_blank()) +
  ggtitle("V0") +
  scale_x_date(date_labels = "'%y") +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))

gg_m1_V1 <- ggplot() +
  geom_line(data = h1_S2_results$mod1_V1_sims, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_ribbon(data = h1_S2_results$mod1_V1_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  ylab("Log reported cases") +
  theme(axis.title = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  ggtitle("V1") +
  scale_x_date(date_labels = "'%y") +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))

gg_m1_V2 <- ggplot() +
  geom_line(data = h1_S4_results$mod1_V2_sims, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_ribbon(data = h1_S4_results$mod1_V2_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  ylab("Log reported cases") +
    theme(axis.title = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  ggtitle("V2") +
  scale_x_date(date_labels = "'%y") +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))

gg_m1_V3 <- ggplot() +
  geom_line(data = h1_S3_results$mod1_V3_sims, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_ribbon(data = h1_S3_results$mod1_V3_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  ylab("Log reported cases") +
    theme(axis.title = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  ggtitle("V3") +
  scale_x_date(date_labels = "'%y") +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))

gg_m1_V4 <- ggplot() +
  geom_line(data = h1_S1_results$mod1_V4_sims, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_ribbon(data = h1_S1_results$mod1_V4_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  ylab("Log reported cases") +
    theme(axis.title = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  ggtitle("V4") +
  scale_x_date(date_labels = "'%y") +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))

\begin{figure}
<<Plot_Model1_Scenarios, fig.height=2>>=
lay <- rbind(c(1, 1, 1, 1, 2, 2, 2, 3, 3, 3, 4, 4, 4, 5, 5, 5))
gridExtra::grid.arrange(
  gg_m1_V0, gg_m1_V1, gg_m1_V2, gg_m1_V3, gg_m1_V4,
  layout_matrix = lay
)
@
\caption{\label{fig:Mod1Scenarios}
Simulations of Model~1 under each vaccination scenario.
Blue line indicates the simulated median of reported cases, and the ribbon represents $95\%$ of simulations.
% The various vaccination campaigns made no practical difference in the median scenario, but a drastic difference in the extreme cases.
}
\end{figure}

\begin{figure}
<<Mod2Fit_and_Scenarios_Figure, fig.height=1.9, fig.width=4.8, fig.align='center'>>=
ggplot() +
  geom_line(data = h2_traj, aes(x = date, y = Ctotal + 1), col = 'blue') +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  geom_line(data = filter(mod2_VaccScenarios$all_trajs,
                          year >= max(h2_traj$year)),
            aes(x = as.Date(date), y = ReportedCases + 1, color = scenario)) +
  theme(axis.title.x = element_blank(),
        legend.title = element_blank(),
        axis.title.y = element_text(size = 9),
        axis.text = element_text(size = 8)) +
  ylab("Reported cholera cases") +
  scale_color_manual(values = c("V0" = '#f4a582',
                                "V1" = '#d6604d',
                                "V2" = '#b2182b',
                                "V3" = '#92c5de',
                                "V4" = '#4393c3')) +
  # geom_vline(xintercept = yearsToDate(2014.161), linetype = 'dashed') +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2010-01-01"), as.Date("2030-01-01"), by = '2 years'))
@
\caption{\label{fig:mod2Traj}
Simulated trajectory of Model~2 (blue curve) and projections under the various vaccination scenarios.
Reported cholera incidence is shown in black.}
\end{figure}

<<Create Model 3 VaccScen Plots>>=
gg_m3_V0 <- ggplot() +
  geom_line(data = mod3_V0_res$mod3_V0_sims, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_ribbon(data = mod3_V0_res$mod3_V0_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  ylab("Log reported cases") +
  theme(axis.title.x = element_blank()) +
  ggtitle("V0") +
  scale_x_date(date_labels = "'%y") +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))

gg_m3_V1 <- ggplot() +
  geom_line(data = mod3_V1_res$mod3_V1_sims, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_ribbon(data = mod3_V1_res$mod3_V1_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  ylab("Log reported cases") +
  theme(axis.title = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  ggtitle("V1") +
  scale_x_date(date_labels = "'%y") +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))

gg_m3_V2 <- ggplot() +
  geom_line(data = mod3_V2_res$mod3_V2_sims, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_ribbon(data = mod3_V2_res$mod3_V2_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  ylab("Log reported cases") +
    theme(axis.title = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  ggtitle("V2") +
  scale_x_date(date_labels = "'%y") +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))

gg_m3_V3 <- ggplot() +
  geom_line(data = mod3_V3_res$mod3_V3_sims, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_ribbon(data = mod3_V3_res$mod3_V3_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  ylab("Log reported cases") +
    theme(axis.title = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  ggtitle("V3") +
  scale_x_date(date_labels = "'%y") +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))

gg_m3_V4 <- ggplot() +
  geom_line(data = mod3_V4_res$mod3_V4_sims, aes(x = date, y = q50 + 1), col = 'blue') +
  geom_ribbon(data = mod3_V4_res$mod3_V4_sims, aes(x = date, ymin = q05 + 1, ymax = q95 + 1), alpha = 0.5) +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  ylab("Log reported cases") +
    theme(axis.title = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  ggtitle("V4") +
  scale_x_date(date_labels = "'%y") +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x)))
@

\begin{figure}
<<Plot_Model3_Scenarios, fig.height=2>>=
lay <- rbind(c(1, 1, 1, 1, 2, 2, 2, 3, 3, 3, 4, 4, 4, 5, 5, 5))
gridExtra::grid.arrange(
  gg_m3_V0, gg_m3_V1, gg_m3_V2, gg_m3_V3, gg_m3_V4,
  layout_matrix = lay
)
@
\caption{\label{fig:Mod3Scenarios}
Simulations of Model~3 under each vaccination scenario.
Blue line indicates the simulated median of reported cases, and ribbon represents $95\%$ of simulations.
% The various vaccination campaigns made no practical difference in the median scenario, but a drastic difference in the extreme cases.
}
\end{figure}

@@

\begin{figure}
<<Elimination_Probs_Figure, fig.height=2.2>>=
ggplot(tidyr::drop_na(all_prob_elims), aes(x = time, y = elim_prob, col = mod)) +
  geom_line() +
  facet_wrap(~scenario, nrow = 1) +
  scale_color_manual(values = c("Model 1" = "#377eb8", "Model 3" = "#e41a1c")) +
  ylab("Probability of Elimination (%)") +
  theme(axis.title.x = element_blank(),
        legend.position = 'bottom',
        legend.title = element_blank(),
        axis.text.x = element_text(angle = 35, hjust = 1),
        legend.margin=margin(c(-5, 0, -3, 0)),
        legend.box.margin = margin(c(-5, 0, -3, 0))) +
  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2010-01-01"), as.Date("2030-01-01"), by = '2 years'))
@
\caption{\label{fig:elimProbs}
Probability of elimination across simulations for a 10 year period.
Compare to Fig.~3A of \cite{lee20}.}
\end{figure}

@@

  & \multicolumn{1}{c}{\hspace{16pt}\missing}

@@

  & ${\fixed \Wremoval^{-1} = \Sexpr{signif(52/h2_fit$h2_params["Delta"],3)}}$
  & \eqref{model2:from_W}
  & $\Wremoval^{-1}=\Sexpr{myround(52/p3s["mu_B"],2)}$
  & \eqref{eq:model3:Decay}

@@

Process noise \newline ($\text{wk}^{1/2}$)

@@

Observation variance

@@

\end{tabular}
\caption{References to the relevant equation are given in parentheses.
Parameters in blue were fixed based on scientific reasoning and not fitted to the data.
$^*$ denotes parameters added during our re-analysis, not considered by Lee et al.
Translations back into the notation of \cite{lee20} are given in Table~S1.
}
\end{adjustwidth}
\end{table}

% Place tables after the first paragraph in which they are cited.
\begin{table}[!ht]
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\centering
\caption{
{\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
\begin{tabular}{|l+l|l|l|l|l|l|l|}
\hline
\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \thickhline
$cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
$cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
$cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
\end{tabular}
\begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl. Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
\end{flushleft}
\label{table1}
\end{adjustwidth}
\end{table}

>>>>

This case study investigates the use of dynamic models and spatiotemporal data to inform a policy decision in the context of the cholera outbreak in Haiti, which started in 2010.
We build on a multi-group modeling exercise by Lee et al.~\cite{lee20} in which four expert modeling teams developed models to the same dataset with the goal of comparing conclusions on the feasibility of eliminating cholera by a vaccination campaign.
The data that was used for this study, and that is the subject of our reanalysis, is displayed in Fig.~\ref{fig:CholeraData}.
Model~1 is stochastic and describes cholera at the national level;
Model~2 is deterministic with spatial structure, and includes transmission via contaminated water;
Model~3 is stochastic with spatial structure, and accounts for measured rainfall.
Model~4 has an agent-based construction, featuring considerable mechanistic detail but limited ability to calibrate these details to data.
We focus on Models~1--3, as the strengths and weaknesses of the agent-based modeling approach \cite{tracy18} are outside the scope of this article.

\begin{figure}[!h]
<<Plot_Reported_Cases, echo=FALSE, fig.height=3.5>>=

# Plot reported cholera cases, by department. Creates Figure 1.
ggplot(dep_plot_df, aes(x = date, y = cases + 1)) +
  facet_wrap(~dep, nrow = 2, labeller = dep_labeller) +
  geom_line() +
  theme(axis.title.x = element_blank()) +
  ylab('Reported Cases') +
  scale_y_log10(
    labels = scales::trans_format("log10", scales::math_format(10^.x))
  ) +
  scale_x_date(
    date_labels = "'%y",
    breaks = seq.Date(
      from = as.Date("2011-01-01"),
      to = as.Date("2019-01-01"),
      by = '2 years'
    )
  )
@
\caption{\label{fig:CholeraData}
\small
Weekly reported cholera cases in Haiti from October 2010 to January 2019.
}
\end{figure}

@@

@@

One of the most frequently used approaches to calibrate the parameters of a statistical model is maximum likelihood estimation.
In the case of a POMP model, the likelihood function---as described in Eq.~\myeqref{eq:likedef}---is written as $\mathcal{L}(\paramVec) = \int f_{\bm{X}_{0:N}, \bm{Y}_{1:N}}\big(\bm{x}_{0:N}, \bm{y}_{1:N};\paramVec\big) d\bm{x}_{0:N}$.
The high-dimensional integral in the likelihood, and the potentially non-linear nature of the process model, make it impossible to obtain a closed form expression of the likelihood function of POMP models except in the most simple cases.
The process model that defines Model~1 was developed with the intent of providing a plausible description of cholera dynamics rather than a providing statistically convenient description.
Because of this, a closed form expression of the likelihood of Model~1 is not readily available.

@@

The ability to maximize the likelihood allows for likelihood-based inference, such as performing statistical tests for potential model improvements.
We demonstrate this capability by proposing a linear trend $\transmissionTrend$ in transmission in Eq.~\myeqref{model1:beta}:
\begin{equation}
\label{model1:betat}
\log\transmission(t) = \sum_{j=1}^6 \transmission_s s_j(t) + \transmissionTrend\bar{t},
\end{equation}
where $\bar{t} = \frac{t - (t_N + t_0) / 2}{t_N - (t_N + t_0) / 2}$, so that $\bar{t} \in [-1, 1]$.
The proposal of a linear trend in transmission is a result of observing an apparent decrease in reported cholera infections from 2012-2019 in Fig.~\ref{fig:CholeraData}.
While several factors may contribute to this decrease, one explanation is that case-area targeted interventions (CATIs), which included education sessions, increased monitoring, household decontamination, soap distribution, and water chlorination in infected areas \cite{rebaudet19CATI}, may have substantially reduced cholera transmission over time \cite{rebaudet21}.

@@

\begin{figure}[!h]
\centering
<<Beta_trend_Figure, fig.height=2.4, fig.width=3.6, fig.align='center'>>=
ggplot() +
  geom_point(data = h1_prof_res, aes(x = betat, y = logLik)) +
  geom_line(data = mcap_results$fit, aes(x = parameter, y = smoothed), col = 'blue') +
  geom_vline(xintercept = mcap_results$ci[1], linetype = 'dashed') +
  geom_vline(xintercept = mcap_results$ci[2], linetype = 'dashed') +
  geom_vline(xintercept = mcap_results$mle, col = 'blue') +
  labs(x = "Linear Trend in Transmission", y = 'Log Likelihood') +
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 10))
@
\caption{\label{fig:betat}
Monte Carlo adjusted profile of $\transmissionTrend$.
The blue curve is the profile, the blue line indicates the MLE, and the dashed lines indicate the $95\%$ confidence interval.
}
\end{figure}

We implemented Model~1 using the \code{pomp} package \cite{king16}, relying heavily on the \code{pomp} source code provided by Lee et al.~\cite{lee20}.
Both analyses used the \code{mif2} implementation of the IF2 algorithm to estimate $\paramVec$ by maximum likelihood.
One change we made in the statistical analysis that led to larger model likelihoods was increasing the computational effort in the numerical maximization.
While IF2 enables parameter estimation for a large class of models, the theoretic ability to maximize the likelihood depends on asymptotics in both the number of particles and the number of filtering iterations.
Many Monte Carlo replications are then required to quantify and further reduce the error.
The large increase in the log-likelihood for Model~1 (see Table~\ref{tab:likes}) can primarily be attributed to increasing the computational effort in fitting the model; this result highlights the importance of carefully determining the necessary computational effort needed to maximize model likelihoods and acting accordingly.
Simulations from the initial conditions of our fitted model are plotted against the observed incidence data in Fig.~\ref{fig:mod1fit}.

\begin{figure}[!h]
\centering
<<Model_1_Sims_Figure, fig.height=2.35, fig.width=4.6, fig.align='center'>>=
# This code chunk creates the figure containing simulations from Model 1

@@

\begin{figure}[!h]
<<Mod2Fit_and_Scenarios_Figure, fig.height=1.9, fig.width=4.8, fig.align='center'>>=
ggplot() +
  geom_line(data = h2_traj, aes(x = date, y = Ctotal + 1), col = 'blue') +
  geom_line(data = true_agg_cases, aes(x = date, y = ReportedAll + 1)) +
  geom_line(data = filter(mod2_VaccScenarios$all_trajs,
                          year >= max(h2_traj$year)),
            aes(x = as.Date(date), y = ReportedCases + 1, color = scenario)) +
  theme(axis.title.x = element_blank(),
        legend.title = element_blank(),
        axis.title.y = element_text(size = 9),
        axis.text = element_text(size = 8)) +
  ylab("Reported cholera cases") +
  scale_color_manual(values = c("V0" = '#f4a582',
                                "V1" = '#d6604d',
                                "V2" = '#b2182b',
                                "V3" = '#92c5de',
                                "V4" = '#4393c3')) +
  # geom_vline(xintercept = yearsToDate(2014.161), linetype = 'dashed') +
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2010-01-01"), as.Date("2030-01-01"), by = '2 years'))
@
\caption{\label{fig:mod2Traj}
Simulated trajectory of Model~2 (blue curve) and projections under the various vaccination scenarios.
Reported cholera incidence is shown in black.}
\end{figure}

@@

At the time Lee et al.~\cite{lee20} was published, there was no known algorithm that could readily be used to maximize the likelihood of an arbitrary meta-population POMP model with coupled spatial dynamics, which justifies the spatial decoupling approximation that was used to calibrate model parameters.
For our analysis, we calibrate the parameters of the spatially coupled version of Model~3 using the iterated block particle filter (IBPF) algorithm \cite{ionides22}.
This algorithm extends the work of Ning and Ionides \cite{ning21ibpf}, who provided theoretic justification for the version of the algorithm that only estimates unit-specific parameters.
The IBPF algorithm enables us to directly estimate the parameters of models describing high-dimensional partially-observed nonlinear dynamic systems via likelihood maximization.
The ability to directly estimate parameters of Model~3 is responsible for the large increase in model likelihoods reported in Table~\ref{tab:likes}.
Simulations from the fitted model are displayed in Fig.~\ref{fig:h3spatsims}.

@@

\begin{figure}[!h]
<<Model3_SpatPOMP_sims_Figure, fig.height=3>>=

@@

<<Get Model 1 Seasonality>>=
h1_seas_df <- get_h1_seasonality(h1)

@@

\begin{figure}[!h]
\centering
<<Model1_Seasonality_Figure, fig.height=2.9, fig.width=5, fig.align='center'>>=
cowplot::plot_grid(gg_rain, gg_trans, align = "v", ncol = 1)
@
\caption{\label{fig:h1SeasRain}
(Top) weekly rainfall in Haiti, lighter colors representing more recent years.
(Bottom) estimated seasonality in the transmission rate (dashed line) plotted alongside mean rainfall (solid line).
The outsided effect of rainfall in the fall may be due to Hurricane Matthew, which struck Haiti in October of 2016 and resulted in an increase of cholera cases in the nation.
}
\end{figure}

@@

Simulations from probabilistic models (Models~1 and~3) represent possible trajectories of the dynamic system under the scientific assumptions of the models.
Estimates of the probability of cholera elimination can therefore be obtained as the proportion of simulations from these models that result in cholera elimination.
The results of these projections are summarized in Figs. \ref{fig:elimProbs}.

@@

\begin{figure}[!h]
<<Elimination_Probs_Figure, fig.height=2.2>>=
ggplot(tidyr::drop_na(all_prob_elims), aes(x = time, y = elim_prob, col = mod)) +
  geom_line() +
  facet_wrap(~scenario, nrow = 1) +
  scale_color_manual(values = c("Model 1" = "#377eb8", "Model 3" = "#e41a1c")) +
  ylab("Probability of Elimination (%)") +
  theme(axis.title.x = element_blank(),
        legend.position = 'bottom',
        legend.title = element_blank(),
        axis.text.x = element_text(angle = 35, hjust = 1),
        legend.margin=margin(c(-5, 0, -3, 0)),
        legend.box.margin = margin(c(-5, 0, -3, 0))) +
  scale_x_date(date_labels = "'%y", breaks = seq.Date(from = as.Date("2010-01-01"), as.Date("2030-01-01"), by = '2 years'))
@
\caption{\label{fig:elimProbs}
Probability of elimination across simulations for a 10 year period.
Compare to Fig.~3A of Lee et al.~\cite{lee20}.}
\end{figure}

@@

@@

  & \multirow{2}{*}{${\fixed \Wremoval^{-1} = \Sexpr{signif(52/h2_fit$h2_params["Delta"],3)}}$}
  & \multirow{2}{*}{\eqref{model2:from_W}}
  & \multirow{2}{*}{$\Wremoval^{-1}=\Sexpr{myround(52/p3s["mu_B"],2)}$}
  & \multirow{2}{*}{\eqref{eq:model3:Decay}}

@@

Process noise \newline (wk\textsuperscript{1/2})

@@

Observation \newline variance

@@

\end{tabular}
\caption{\label{tab:params}References to the relevant equation are given in parentheses.
Parameters in blue were fixed based on scientific reasoning and not fitted to the data.
$^*$ denotes parameters added during our re-analysis, not considered by Lee et al.
Translations back into the notation of \cite{lee20} are given in Table~S1.
}
\end{adjustwidth}
\end{table}

% TODO: Remove the unused code below.
% % Place tables after the first paragraph in which they are cited.
% \begin{table}[!ht]
% \begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
% \centering
% \caption{
% {\bf Table caption Nulla mi mi, venenatis sed ipsum varius, volutpat euismod diam.}}
% \begin{tabular}{|l+l|l|l|l|l|l|l|}
% \hline
% \multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \thickhline
% $cell1 row1$ & cell2 row 1 & cell3 row 1 & cell4 row 1 & cell5 row 1 & cell6 row 1 & cell7 row 1 & cell8 row 1\\ \hline
% $cell1 row2$ & cell2 row 2 & cell3 row 2 & cell4 row 2 & cell5 row 2 & cell6 row 2 & cell7 row 2 & cell8 row 2\\ \hline
% $cell1 row3$ & cell2 row 3 & cell3 row 3 & cell4 row 3 & cell5 row 3 & cell6 row 3 & cell7 row 3 & cell8 row 3\\ \hline
% \end{tabular}
% \begin{flushleft} Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus tortor nec nisl. Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
% \end{flushleft}
% \label{table1}
% \end{adjustwidth}
% \end{table}
